---
title: "Comorbidity Patterns using Hierarchical Clustering"
subtitle: "Unsupervised analysis (Jaccard distance)"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```  
   
# 1. Introduction and Configuration  
  
The objective of this script is to identify complex clusters of diseases (comorbidity patterns) in a unsupervised manner.   
  
Since the diagnostic matrix is composed of binary variables (0/1), the Jaccard distance is used to avoid the bias of shared absences (0-0) and to evaluate similarity based solely on the co-occurrence of pathologies.  
  
# 2. Data Loading and Preparation  
  
We load the prevalence dataset and filter out non-diagnostic variables. We also apply a prevalence threshold of 2%.  
  
```{r}
# Upload prevalent diagnostic data
diag_data <- read_csv(PATH_DIAG_BOOL_PREVALENT_FINAL)

# Filter disease matrix (exclude identifiers and HIV-related variables)
disease_df <- diag_data %>% 
  select(-patient_id, -vih_dt, -HIV_infection, -`HIV-related_disease`) 
 
# Apply a minimum prevalence filter of 2%
min_prev <- 0.02
prevalences <- colMeans(disease_df)
frequent_diseases <- names(prevalences[prevalences >= min_prev])

# Filter matrix keeping only pathologies >= 2%
disease_df_filtered <- disease_df %>% 
  select(all_of(frequent_diseases))

# Convert to numeric matrix for R calculations
matrix_disease <- as.matrix(disease_df_filtered)

dim(disease_df_filtered)
```  
  
# 3. Transposed Matrix and Jaccard Distance  
  
To group variables (diseases) instead of observations (patients), it is necessary to transpose the matrix before calculating the distances. We use the Jaccard distance, which is mathematically robust for binary data as it focuses on the presence of conditions.  
  
$$\text{Jaccard Distance} = 1 - \frac{A \cap B}{A \cup B}$$  
```{r}
# Transpose the matrix: Rows = Diseases | Columns = Patients
diseases_t <- t(matrix_disease)

# Calculate Jaccard distance matrix between pathologies
jaccard_dist <- proxy::dist(diseases_t, method = "Jaccard")
```  
  
## 3.1. Clustering Tendency Assessment  
  
Before applying hierarchical clustering, we must assess whether the data has an inherent structure or is just random noise. We use the Hopkins Statistic and Visual Assessment of Cluster Tendency (VAT). A Hopkins value significantly above 0.7 indicates a high propensity for natural cluster formation.  
  
```{r}
# 1. Project Jaccard distance into reduced Euclidean space (MDS)
mds_coords <- cmdscale(jaccard_dist, k = min(5, nrow(diseases_t) - 1))

# 2. Define a safe sample size for Hopkins
n_sample <- max(5, min(10, nrow(mds_coords) - 1))

# 3. Calculate Hopkins Statistic
hopkins_res <- get_clust_tendency(
  data = mds_coords, 
  n = n_sample, 
  graph = FALSE
)

h_val <- hopkins_res$hopkins_stat

if (!is.null(h_val) && !is.na(h_val)) {
  message("Estadístico de Hopkins (H): ", round(h_val, 3))
  
  if (h_val > 0.7) {
    message("-> CONCLUSION: The data show a significant tendency to cluster.")
  } else {
    message("-> WARNING: Moderate or weak clustering structure.")
  }
} else {
  message("-> WARNING: Could not compute Hopkins statistic.")
}

# 4. VAT plot
fviz_dist(
  jaccard_dist, 
  show_labels = FALSE, 
  gradient = list(low = "red", mid = "white", high = "#3C58D0")
) + 
labs(
  title = "Visual Clustering Tendency Assessment (Jaccard VAT)",
  fill = "Jaccard Dist"
)

print(h_val)
```  
  
The Hopkins statistic ($H \approx 0.815$) confirms that the pathologies are not randomly distributed, justifying the use of hierarchical clustering as a reliable method to extract meaningful comorbidity patterns.  
  
# 4. Hierarchical clustering (Ward.D2)  
  
We performed hierarchical clustering using the Ward.D2 linkage metric, which minimizes variance within the resulting clusters.  
  
```{r}
# Adjust the hierarchical clustering model
hc_diseases <- hclust(jaccard_dist, method = "ward.D2")
```  
  
# 5. Selecting the Optimal Number of Patterns (k)  
  
Instead of assuming a number of patterns a priori or applying generic empirical rules (such as $k = \sqrt{N/2}$), we evaluate the **average silhouette width** on the Jaccard dissimilarity matrix for $k \in [3, 10]$.

```{r optimal-k}
k_range <- 3:10

# Calculate the silhouette profile to assess group cohesion
sil_values <- sapply(k_range, function(k_val) {
  summary(silhouette(cutree(hc_diseases, k = k_val), jaccard_dist))$avg.width
})

# Graphically represent the result
plot(k_range, sil_values, type = "b", pch = 19,
     xlab = "Number of Clusters (k)", 
     ylab = "Average Silhouette",
     main = "Evaluation of k using Average Silhouette Width")

k_optimal <- 5
```  
  
While the silhouette metric mathematically suggested $k = 9$, we chose $k = 5$. Choosing a lower $k$ avoids over-fragmentation (where some clusters contain only one disease) and ensures better clinical interpretability, maintaining consistency with our previous Exploratory Factor Analysis (which identified 4-6 factors).  
  
# 6. Visualization of the Dendrogram  
  
We visualize the hierarchical structure of pathologies to analyze the natural patterns of comorbidity.  
  
```{r}
# Graph the dendrogram of pathologies
dend <- as.dendrogram(hc_diseases)

# Customize size and colors
dend %>% 
  set("labels_cex", 0.6) %>% 
  set("branches_k_color", k = k_optimal) %>% 
  plot(
    main = "Comorbidity Pattern Dendrogram (Jaccard + Ward.D2)",
    ylab = "Height (Jaccard Distance)"
  )
```  
  
# 7. Pattern Extraction and Export  
  
We use the optival k value to extract the lists of pathologies belonging to each pattern.  
  
```{r}
k_clusters <- k_optimal

diseases_clusters <- cutree(hc_diseases, k = k_clusters)

# Disease to cluster mapping
df_patrones <- data.frame(
  enfermedad = names(diseases_clusters),
  cluster = factor(diseases_clusters)
) %>% 
  arrange(cluster)

# Save results in CSV
write_csv(
  df_patrones,
  file.path(PATH_INTERMEDIATE, "jaccard_disease_patterns.csv")
)

# Show summary of pathologies by cluster
df_patrones %>% 
  group_by(cluster) %>% 
  summarise(n_enfermedades = n(), ejemplos = paste(head(enfermedad, 3), collapse = ", "))
```
