---
title: "Comorbidity Patterns using Hierarchical Clustering"
subtitle: "Unsupervised analysis (Jaccard distance)"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```  
  
# 1. Introduction and Configuration  
  
The objective of this script is to identify complex clusters of diseases (comorbidity patterns) in a completely unsupervised manner.   
  
Since the diagnostic matrix is composed of binary variables (0/1), the Jaccard distance is used to avoid the bias of shared absences (0-0) and to evaluate similarity based solely on the co-occurrence of pathologies.  
  
# 2. Data Loading and Preparation  
  
We load the prevalence dataset and prepare the excluded matrix of primary study identifiers and variables (HIV_infection, HIV-related_disease).  
  
```{r}
# Upload prevalent diagnostic data
diag_data <- read_csv(PATH_DIAG_BOOL_PREVALENT_FINAL)

# Filter disease matrix
disease_df <- diag_data %>% 
  select(-patient_id, -vih_dt, -HIV_infection, -`HIV-related_disease`)

# Convert to numeric matrix R
matrix_disease <- as.matrix(disease_df)

dim(matrix_disease)
```  
  
# 3. Transposed Matrix and Jaccard Distance  
  
To group variables (diseases) instead of observations (patients), it is necessary to transpose the matrix before calculating the distances.  
  
$$\text{Jaccard Distance} = 1 - \frac{A \cap B}{A \cup B}$$  
  
```{r}
# Transpose the matrix: Rows = Diseases | Columns = Patients
diseases_t <- t(matrix_disease)

# Calculate Jaccard distance matrix between pathologies
jaccard_dist <- proxy::dist(diseases_t, method = "Jaccard")
```  
  
# 4. Hierarchical clustering (Ward.D2)  
  
We performed hierarchical clustering using the Ward.D2 linkage metric, which minimizes variance within the resulting clusters.  
  
```{r}
# Adjust the hierarchical clustering model
hc_diseases <- hclust(jaccard_dist, method = "ward.D2")
```  
  
# 5. Selecting the Optimal Number of Patterns (k)  
  
Instead of assuming a number of patterns a priori or applying generic empirical rules (such as $k = \sqrt{N/2}$), we evaluate the **average silhouette width** on the Jaccard dissimilarity matrix for $k \in [2, 15]$.

```{r optimal-k}
# Calculate the silhouette profile to assess group cohesion
sil_values <- sapply(2:15, function(k_val) {
  summary(silhouette(cutree(hc_diseases, k = k_val), jaccard_dist))$avg.width
})

# Graphically represent the result
plot(2:15, sil_values, type = "b", pch = 19,
     xlab = "Number of Clusters (k)", 
     ylab = "Average Silhouette",
     main = "Evaluation of k using Average Silhouette Width")

# Automatically select optimal k
k_optimal <- which.max(sil_values) + 1   
```    
  
# 6. Visualization of the Dendrogram  
  
We visualize the hierarchical structure of pathologies to analyze the natural patterns of comorbidity.  
  
```{r}
# Graph the dendrogram of pathologies
dend <- as.dendrogram(hc_diseases)

# Customize size and colors
dend %>% 
  set("labels_cex", 0.5) %>% 
  set("branches_k_color", k = k_optimal) %>% 
  plot(
    main = "Comorbidity Pattern Dendrogram (Jaccard + Ward.D2)",
    ylab = "Height (Jaccard Distance)"
  )
```  
  
# 7. Pattern Extraction and Export  
  
We use the optival k value to extract the lists of pathologies belonging to each pattern.  
  
```{r}
k_clusters <- k_optimal
diseases_clusters <- cutree(hc_diseases, k = k_clusters)

# Disease to cluster mapping
df_patrones <- data.frame(
  enfermedad = names(diseases_clusters),
  cluster = factor(diseases_clusters)
) %>% 
  arrange(cluster)

# Save results in CSV
write_csv(
  df_patrones,
  file.path(PATH_INTERMEDIATE, "jaccard_disease_patterns.csv")
)

# Show summary of pathologies by cluster
df_patrones %>% 
  group_by(cluster) %>% 
  summarise(n_enfermedades = n(), ejemplos = paste(head(enfermedad, 3), collapse = ", "))
```

