---
title: "Comorbidity Patterns using Hierarchical Clustering"
subtitle: "Unsupervised analysis (Jaccard distance)"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8
)
```  
   
# 1. Introduction and Configuration  
  
The objective of this script is to identify clusters of diseases (comorbidity patterns) using an unsupervised hierarchical clustering approach.   
  
Unlike factor-based methods, hierarchical clustering does not assume that
the observed disease variables arise from a predefined latent-factor
structure. Instead, diseases are grouped according to their similarity in
co-occurrence patterns.

Because the diagnostic matrix is binary (0/1), the Jaccard distance is used as the primary dissimilarity measure. The Jaccard distance focuses on shared disease presence and ignores shared absences (0-0), which is appropriate when the absence of a disease is not considered informative for defining multimorbidity patterns.

The analysis consists of the following main steps:

1. Preparation of the disease prevalence matrix 
2. Exploratory assessment of binary disease associations
3. Calculation of Jaccard dissimilarities between diseases 
4. Assessment of clustering tendency 
5. Hierarchical clustering using the Ward.D2 criterion
6. Evaluation of candidate solutions with 3 to 7 clusters
7. Extraction and export of disease-to-cluster assignments  
  
# 2. Data Loading and Preparation  
  
We load the prevalence dataset and filter out non-diagnostic variables. We also apply a prevalence threshold of 2%.  
  
```{r}
# Load prevalent diagnostic data
diag_data <- read_csv(PATH_DIAG_BOOL_PREVALENT_FINAL)

# Remove patient identifiers and HIV-related variables that are not used
# as disease indicators in the clustering analysis
disease_df <- diag_data %>% 
  select(-patient_id, -vih_dt, -HIV_infection, -`HIV-related_disease`) 
 
# Apply a minimum prevalence filter of 2%
min_prev <- 0.02

# Calculate the prevalence of each disease across the study population
prevalences <- colMeans(disease_df)

# Identify diseases meeting the predefined 2% prevalence threshold
frequent_diseases <- names(prevalences[prevalences >= min_prev])

# Filter matrix keeping only pathologies >= 2%
disease_df_filtered <- disease_df %>% 
  select(all_of(frequent_diseases))

# Convert to numeric matrix for R calculations
matrix_disease <- as.matrix(disease_df_filtered)

dim(disease_df_filtered)
```  
  
# 3. Binary Disease Correlation Matrix

As an exploratory analysis, a Pearson correlation matrix is calculated for
the binary disease indicators.  
  
The diseases are subsequently reordered according to their similarity in
correlation profiles. This reordering is performed exclusively to improve
the visual organisation of the heatmap and does not define the final
Jaccard-based disease clusters.

A square-root transformation is applied only to the values used to define
the colour scale. This transformation increases the visual separation of
small correlation values while preserving their direction. The original
correlation values are retained and displayed in the interactive tooltips.

```{r}
# Tetrachoric correlation matrix for binary data
corr_matrix <- psych::tetrachoric(matrix_disease,na.rm = TRUE)$rho

# Cleanse disease names
clean_names <- gsub("_", " ", colnames(matrix_disease))
rownames(corr_matrix) <- clean_names
colnames(corr_matrix) <- clean_names

# Hierarchical clustering is used only to reorder diseases in the
# correlation heatmap for visualisation. It is not used to define
# the Jaccard-based disease patterns
row_dist <- as.dist(1 - corr_matrix)
row_clust <- hclust(row_dist, method = "ward.D2") 

# Extract the order generated by the hierarchical clustering
ordered_names <- row_clust$labels[row_clust$order]
# Reorder rows and columns using the same disease order
corr_ordered <- corr_matrix[ordered_names, ordered_names]

# Remove the diagonal because self-correlations are not informative
# for visualising relationships between different diseases
diag(corr_ordered) <- NA

zmin <- -0.15
zmax <-  0.15

breaks <- c(
  -0.15,
  -0.125,
  -0.10,
  -0.075,
  -0.05,
  -0.025,
   0,
   0.025,
   0.05,
   0.075,
   0.10,
   0.125,
   0.15
)

colors <- c(
  "#67000D",
  "#A50F15",
  "#CB181D",
  "#EF3B2C",
  "#FC8D59",
  "#FFFFFF",
  "#FFFFFF",
  "#9ECAE1",
  "#6BAED6",
  "#4292C6",
  "#2171B5",
  "#08519C"
)

# Build a truly DISCRETE Plotly colour scale
colorscale <- list()

for (i in seq_len(length(colors))) {
  
  lower <- breaks[i]
  upper <- breaks[i + 1]
  
  lower_norm <- (lower - zmin) / (zmax - zmin)
  upper_norm <- (upper - zmin) / (zmax - zmin)
  
  # Same colour at both ends of the interval
  colorscale[[length(colorscale) + 1]] <- 
    list(lower_norm, colors[i])
  
  colorscale[[length(colorscale) + 1]] <- 
    list(upper_norm, colors[i])
}

# Interactive heatmap
plot_ly(
  x = colnames(corr_ordered),
  y = rownames(corr_ordered),
  z = corr_ordered,
  type = "heatmap",
  colorscale = colorscale,
  zmin = zmin,
  zmax = zmax,
  hoverongaps = FALSE,
  text = round(corr_ordered, 3),
  hovertemplate = paste0(
    "Disease 1: %{y}<br>",
    "Disease 2: %{x}<br>",
    "Tetrachoric correlation: %{text}<extra></extra>"
  ),
  colorbar = list(
    title = "Tetrachoric correlation",
    tickmode = "array",
    tickvals = breaks,
    ticktext = sprintf("%.2f", breaks)
  )
) %>%
  layout(
    title = "Binary Tetrachoric Correlation Matrix",
    xaxis = list(
      title = "",
      tickangle = -45,
      tickfont = list(size = 8),
      categoryarray = colnames(corr_ordered),
      categoryorder = "array"
    ),
    yaxis = list(
      title = "",
      tickfont = list(size = 8),
      autorange = "reversed",
      categoryarray = rownames(corr_ordered),
      categoryorder = "array"
    ),
    margin = list(l = 150, b = 150)
  )
```

# 4. Transposed Matrix and Jaccard Distance  
  
The objective of the clustering analysis is to group diseases rather than
patients. Therefore, the disease matrix is transposed so that each disease
becomes an observation and each patient becomes a binary feature.

The Jaccard distance is then calculated between diseases. This metric is
particularly appropriate for binary disease data because it measures
dissimilarity based on shared disease presence while ignoring joint
absences.  
  
$$\text{Jaccard Distance} = 1 - \frac{A \cap B}{A \cup B}$$  
  
where $A \cap B$ represents the number of patients in whom both diseases
are present and $A \cup B$ represents the number of patients in whom at
least one of the two diseases is present.  
  
```{r}
# Transpose the matrix: Rows = Diseases | Columns = Patients
diseases_t <- t(matrix_disease)

# Calculate Jaccard distance matrix between pathologies
jaccard_dist <- proxy::dist(diseases_t, method = "Jaccard")
```  
  
## 4.1. Clustering Tendency Assessment  
  
Before applying hierarchical clustering, the tendency of the disease data
to form groups is explored using the Hopkins statistic.

The Hopkins statistic evaluates whether the observed data exhibit a
clustering tendency that differs from a random spatial distribution.

Because the original Jaccard distance matrix is not directly Euclidean,
the Jaccard dissimilarities are first projected into a reduced Euclidean
space using multidimensional scaling (MDS). The Hopkins statistic is then
calculated on these MDS coordinates.

This analysis is considered an exploratory diagnostic rather than a formal
criterion for selecting the number of clusters. The candidate cluster
solutions are evaluated separately using the average silhouette width.

A VAT (Visual Assessment of cluster Tendency) plot is also generated to
provide a visual assessment of the structure contained in the Jaccard
dissimilarity matrix.  
  
```{r}
# Project Jaccard distance into reduced Euclidean space 
# using multidimensional scaling (MDS)
mds_coords <- cmdscale(jaccard_dist, k = min(5, nrow(diseases_t) - 1))

# Define a safe sample size for Hopkins
n_sample <- max(5, min(10, nrow(mds_coords) - 1))

# Calculate Hopkins Statistic
hopkins_res <- get_clust_tendency(
  data = mds_coords, 
  n = n_sample, 
  graph = FALSE
)

h_val <- hopkins_res$hopkins_stat

# VAT matrix

# Convert Jaccard distance object to matrix
vat_matrix <- as.matrix(jaccard_dist)

# Reorder objects using hierarchical clustering
vat_clust <- hclust(jaccard_dist,method = "ward.D2")
vat_order <- vat_clust$order

# Reorder rows and columns
vat_matrix_ordered <- vat_matrix[vat_order,vat_order]

# Object names
vat_names <- rownames(vat_matrix_ordered)

# Set diagonal to NA so it appears blank in the VAT
diag(vat_matrix_ordered) <- NA

# VAT plot
plot_ly(
  x = vat_names,
  y = vat_names,
  z = vat_matrix_ordered,
  type = "heatmap",
  colorscale = list(
  list(0.00, "#D73027"),  # Lowest distance
  list(0.125, "#D73027"),
  list(0.125, "#F46D43"),
  list(0.25, "#F46D43"),
  list(0.25, "#FDAE61"),
  list(0.375, "#FDAE61"),
  list(0.375, "#FEE090"),
  list(0.50, "#FEE090"),
  list(0.50, "#FFFFBF"),
  list(0.625, "#FFFFBF"),
  list(0.625, "#ABD9E9"),
  list(0.75, "#ABD9E9"),
  list(0.75, "#74ADD1"),
  list(0.875, "#74ADD1"),
  list(0.875, "#4575B4"),
  list(1.00, "#313695")   # Highest distance
),
  zmin = 0.90,
  zmax = 0.995,
  hoverongaps = FALSE,
  text = round(vat_matrix_ordered, 3),
  hovertemplate = paste0(
    "Object 1: %{y}<br>",
    "Object 2: %{x}<br>",
    "Jaccard distance: %{text}<extra></extra>"
  ),
  colorbar = list(
    title = "Jaccard distance",
    tickvals = c(
      0.90,
      0.92,
      0.94,
      0.96,
      0.98,
      0.99,
      0.995
    ),
    ticktext = c(
      "0.90",
      "0.92",
      "0.94",
      "0.96",
      "0.98",
      "0.99",
      "0.995"
    )
  )
) %>%
  layout(
    title = "Visual Clustering Tendency Assessment (Jaccard VAT)",
    xaxis = list(
      title = "",
      tickangle = -45,
      tickfont = list(size = 8),
      categoryarray = vat_names,
      categoryorder = "array"
    ),
    yaxis = list(
      title = "",
      tickfont = list(size = 8),
      autorange = "reversed",
      categoryarray = vat_names,
      categoryorder = "array"
    ),
    margin = list(
      l = 100,
      b = 150
    )
  )

# Display Hopkins Statistic
print(h_val)
```  
  
# 5. Hierarchical clustering (ward.D2 method)  
  
Hierarchical clustering was performed using Ward.D2 on the Jaccard dissimilarity matrix.  
  
```{r}
# Adjust the hierarchical clustering model
hc_diseases <- hclust(jaccard_dist, method = "ward.D2")
```  
  
# 6. Evaluation of Candidate Numbers of Clusters  
  
Candidate hierarchical clustering solutions containing between 3 and 7
clusters are evaluated.

For each candidate value of k, the average silhouette width is calculated
using the original Jaccard dissimilarity matrix. The silhouette width
provides an assessment of how well each disease fits within its assigned
cluster relative to neighbouring clusters.

The silhouette profile is used to characterise the clustering structure
across the candidate values of k. It is not used as the sole criterion to
select a single definitive solution. Instead, the solutions from k = 3 to
k = 7 are retained for subsequent comparison and downstream analyses.

A reference solution with k = 5 is used for the main dendrogram
visualisation and for the corresponding disease-to-cluster mapping.

```{r}
# Define the candidate numbers of clusters
k_range <- 3:7

# Calculate the average silhouette width for each candidate solution
sil_values <- sapply(k_range, function(k_val) {
  summary(silhouette(cutree(hc_diseases, k = k_val), jaccard_dist))$avg.width
})

# Graphically represent the result
plot(k_range, sil_values, type = "b", pch = 19,
     xlab = "Number of Clusters (k)", 
     ylab = "Average Silhouette",
     main = "Evaluation of k using Average Silhouette Width")

# Define the reference solution used for the main visualisation
k_reference <- 5
```  

  
# 7. Visualization of the Dendrogram  
  
The hierarchical clustering dendrogram is visualised to examine the
relationships between diseases and the resulting hierarchical structure.

Branches are coloured according to the reference five-cluster solution to
facilitate visual identification of the corresponding disease groups. 


```{r}
# Piecewise transformation of the dendrogram heights
# Compress the lower range and expand the upper range
transform_y <- function(y) {
  ifelse(
    y <= 0.90,
    (y - 0.87) * 0.25,
    
    ifelse(
      y <= 1.03,
      0.03 * 0.25 + (y - 0.90) * 1.5,
      
      0.03 * 0.25 +
        (1.03 - 0.90) * 1.5 +
        (y - 1.03) * 0.20
    )
  )
}

# Transform dendrogram heights
hc_diseases_trans <- hc_diseases
hc_diseases_trans$height <- transform_y(hc_diseases$height)

# Convert to dendrogram
dend5 <- as.dendrogram(hc_diseases_trans)

# Acortar branches
dend5 <- dendextend::set(
  dend5,
  "hang",
  0.2
)

# Customize label size
dend5 <- dendextend::set(
  dend5,
  "labels_cex",
  0.6
)

# Make branches thicker
dend5 <- dendextend::set(
  dend5,
  "branches_lwd",
  3
)

# Color branches according to the reference solution
dend5 <- dendextend::color_branches(
  dend5,
  k = k_reference
)

# Original values to show on the Y-axis
y_breaks <- c(
  0.90, 0.95, 0.98, 1.00
)

# Increase left margin to prevent disease labels from being cut off
par(mar = c(12, 10, 4, 2) + 0.1)

# Plot dendrogram
plot(
  dend5,
  main = "Comorbidity Pattern Dendrogram (Jaccard + ward.D2)",
  ylab = "",
  yaxt = "n"
)

# Add transformed Y-axis using original heights
axis(
  side = 2,
  at = transform_y(y_breaks),
  labels = sprintf("%.2f", y_breaks),
  las = 1
)

# Y-axis title
mtext(
  "Height (Jaccard Distance)",
  side = 2,
  line = 3
)
```  
  
## 7.1 Visualization for k = 3  
  
```{r}
# Piecewise transformation of the dendrogram heights
# Compress the lower range and expand the upper range
transform_y <- function(y) {
  ifelse(
    y <= 0.90,
    (y - 0.87) * 0.25,
    
    ifelse(
      y <= 1.03,
      0.03 * 0.25 + (y - 0.90) * 1.5,
      
      0.03 * 0.25 +
        (1.03 - 0.90) * 1.5 +
        (y - 1.03) * 0.20
    )
  )
}

# Transform dendrogram heights
hc_diseases_trans <- hc_diseases
hc_diseases_trans$height <- transform_y(hc_diseases$height)

# Convert to dendrogram
dend3 <- as.dendrogram(hc_diseases_trans)

# Acortar branches
dend3 <- dendextend::set(
  dend3,
  "hang",
  0.2
)

# Customize label size
dend3 <- dendextend::set(
  dend3,
  "labels_cex",
  0.6
)

# Make branches thicker
dend3 <- dendextend::set(
  dend3,
  "branches_lwd",
  3
)

# Color branches according to the reference solution
dend3 <- dendextend::color_branches(
  dend3,
  k = 3
)

# Original values to show on the Y-axis
y_breaks <- c(
  0.90, 0.95, 0.98, 1.00
)

# Increase left margin to prevent disease labels from being cut off
par(mar = c(12, 10, 4, 2) + 0.1)

# Plot dendrogram
plot(
  dend3,
  main = "Comorbidity Pattern Dendrogram (Jaccard + ward.D2)",
  ylab = "",
  yaxt = "n"
)

# Add transformed Y-axis using original heights
axis(
  side = 2,
  at = transform_y(y_breaks),
  labels = sprintf("%.2f", y_breaks),
  las = 1
)

# Y-axis title
mtext(
  "Height (Jaccard Distance)",
  side = 2,
  line = 3
)
```   
  
## 7.2 Visualization for k = 4  
  
```{r}
# Piecewise transformation of the dendrogram heights
# Compress the lower range and expand the upper range
transform_y <- function(y) {
  ifelse(
    y <= 0.90,
    (y - 0.87) * 0.25,
    
    ifelse(
      y <= 1.03,
      0.03 * 0.25 + (y - 0.90) * 1.5,
      
      0.03 * 0.25 +
        (1.03 - 0.90) * 1.5 +
        (y - 1.03) * 0.20
    )
  )
}

# Transform dendrogram heights
hc_diseases_trans <- hc_diseases
hc_diseases_trans$height <- transform_y(hc_diseases$height)

# Convert to dendrogram
dend4 <- as.dendrogram(hc_diseases_trans)

# Acortar branches
dend4 <- dendextend::set(
  dend4,
  "hang",
  0.2
)

# Customize label size
dend4 <- dendextend::set(
  dend4,
  "labels_cex",
  0.6
)

# Make branches thicker
dend4 <- dendextend::set(
  dend4,
  "branches_lwd",
  3
)

# Color branches according to the reference solution
dend4 <- dendextend::color_branches(
  dend4,
  k = 4
)

# Original values to show on the Y-axis
y_breaks <- c(
  0.90, 0.95, 0.98, 1.00
)

# Increase left margin to prevent disease labels from being cut off
par(mar = c(12, 10, 4, 2) + 0.1)

# Plot dendrogram
plot(
  dend4,
  main = "Comorbidity Pattern Dendrogram (Jaccard + ward.D2)",
  ylab = "",
  yaxt = "n"
)

# Add transformed Y-axis using original heights
axis(
  side = 2,
  at = transform_y(y_breaks),
  labels = sprintf("%.2f", y_breaks),
  las = 1
)

# Y-axis title
mtext(
  "Height (Jaccard Distance)",
  side = 2,
  line = 3
)
```   
  
## 7.3 Visualization for k = 6 
  
```{r}
# Piecewise transformation of the dendrogram heights
# Compress the lower range and expand the upper range
transform_y <- function(y) {
  ifelse(
    y <= 0.90,
    (y - 0.87) * 0.25,
    
    ifelse(
      y <= 1.03,
      0.03 * 0.25 + (y - 0.90) * 1.5,
      
      0.03 * 0.25 +
        (1.03 - 0.90) * 1.5 +
        (y - 1.03) * 0.20
    )
  )
}

# Transform dendrogram heights
hc_diseases_trans <- hc_diseases
hc_diseases_trans$height <- transform_y(hc_diseases$height)

# Convert to dendrogram
dend6 <- as.dendrogram(hc_diseases_trans)

# Acortar branches
dend6 <- dendextend::set(
  dend6,
  "hang",
  0.2
)

# Customize label size
dend6 <- dendextend::set(
  dend6,
  "labels_cex",
  0.6
)

# Make branches thicker
dend6 <- dendextend::set(
  dend6,
  "branches_lwd",
  3
)

# Color branches according to the reference solution
dend6 <- dendextend::color_branches(
  dend6,
  k = 6
)

# Original values to show on the Y-axis
y_breaks <- c(
  0.90, 0.95, 0.98, 1.00
)

# Increase left margin to prevent disease labels from being cut off
par(mar = c(12, 10, 4, 2) + 0.1)

# Plot dendrogram
plot(
  dend6,
  main = "Comorbidity Pattern Dendrogram (Jaccard + ward.D2)",
  ylab = "",
  yaxt = "n"
)

# Add transformed Y-axis using original heights
axis(
  side = 2,
  at = transform_y(y_breaks),
  labels = sprintf("%.2f", y_breaks),
  las = 1
)

# Y-axis title
mtext(
  "Height (Jaccard Distance)",
  side = 2,
  line = 3
)
```   
  
## 7.4 Visualization for k = 7  
  
```{r}
# Piecewise transformation of the dendrogram heights
# Compress the lower range and expand the upper range
transform_y <- function(y) {
  ifelse(
    y <= 0.90,
    (y - 0.87) * 0.25,
    
    ifelse(
      y <= 1.03,
      0.03 * 0.25 + (y - 0.90) * 1.5,
      
      0.03 * 0.25 +
        (1.03 - 0.90) * 1.5 +
        (y - 1.03) * 0.20
    )
  )
}

# Transform dendrogram heights
hc_diseases_trans <- hc_diseases
hc_diseases_trans$height <- transform_y(hc_diseases$height)

# Convert to dendrogram
dend7 <- as.dendrogram(hc_diseases_trans)

# Acortar branches
dend7 <- dendextend::set(
  dend7,
  "hang",
  0.2
)

# Customize label size
dend7 <- dendextend::set(
  dend7,
  "labels_cex",
  0.6
)

# Make branches thicker
dend7 <- dendextend::set(
  dend7,
  "branches_lwd",
  3
)

# Color branches according to the reference solution
dend7 <- dendextend::color_branches(
  dend7,
  k = 7
)

# Original values to show on the Y-axis
y_breaks <- c(
  0.90, 0.95, 0.98, 1.00
)

# Increase left margin to prevent disease labels from being cut off
par(mar = c(12, 10, 4, 2) + 0.1)

# Plot dendrogram
plot(
  dend7,
  main = "Comorbidity Pattern Dendrogram (Jaccard + ward.D2)",
  ylab = "",
  yaxt = "n"
)

# Add transformed Y-axis using original heights
axis(
  side = 2,
  at = transform_y(y_breaks),
  labels = sprintf("%.2f", y_breaks),
  las = 1
)

# Y-axis title
mtext(
  "Height (Jaccard Distance)",
  side = 2,
  line = 3
)
```   
  
# 8. Reference Pattern Extraction and Export  
  
A reference solution with k = 5 is extracted for visualisation and
inspection. All candidate solutions from k = 3 to k = 7 are exported
separately in Section 9 for downstream analyses.  
  
```{r}
# Set the reference number of clusters
k_clusters <- k_reference

# Assign each disease to one of the reference clusters
diseases_clusters <- cutree(hc_diseases, k = k_clusters)

# Disease to cluster mapping
df_patrones <- data.frame(
  enfermedad = names(diseases_clusters),
  cluster = as.integer(diseases_clusters)
) %>% 
  arrange(cluster)

# Save results in CSV
write_csv(
  df_patrones,
  file.path(PATH_INTERMEDIATE, "jaccard_disease_patterns.csv")
)

# Show summary of pathologies by cluster
tabla_patrones <- df_patrones %>% 
  group_by(cluster) %>% 
  summarise(
    n_enfermedades = n(),
    ejemplos = paste(head(enfermedad, 3), collapse = ", "),
    .groups = "drop"
  ) %>% 
  rename(
    `Cluster` = cluster,
    `Nº de enfermedades` = n_enfermedades,
    `Ejemplos de enfermedades` = ejemplos
  )

tabla_patrones %>% 
  knitr::kable(
    format = "html",
    align = c("c", "c", "l"),
    caption = "Resumen de patologías por clúster"
  ) %>% 
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
```  
  
# 9. Candidate solutions for different numbers of clusters  
  
To support comparison with alternative multimorbidity pattern solutions,
disease partitions are extracted for k = 3 to 7. These candidate solutions
are retained for downstream patient-level analyses and for comparison
across different levels of clustering granularity.  
  
```{r}
# Extract candidate solutions for each k
k_candidatos <- c(3, 4, 5, 6, 7)

# We create a list with a clean data frame for each k using the cutree names
multi_k_list <- map(k_candidatos, function(k_val) {
  cl <- cutree(hc_diseases, k = k_val)
  data.frame(
    enfermedad = names(cl),
    cluster = as.integer(cl)
  ) %>% 
    rename(!!paste0("cluster_k", k_val) := cluster)
})

# We join all the data frames by the 'disease' column without risk of misalignment
df_multi_patrones <- reduce(multi_k_list, full_join, by = "enfermedad")

# Save the multisolution table for clinical review
write_csv(
  df_multi_patrones,
  file.path(PATH_INTERMEDIATE, "jaccard_disease_patterns_multi_k.csv")
)

```
