---
title: "Patient Pattern Scoring"
subtitle: "Mapping disease clusters onto individual patient profiles"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

```

# 1. Load Input Datasets

We import:

1. The binary diagnosis matrix (`diagnosis_bool_prevalent_final.csv`).
2. The mapped disease patterns derived from the unsupervised clustering step (`jaccard_disease_patterns.csv`).

```{r load-data}
# 1. Load patient diagnostic matrix
diag_data <- read_csv(PATH_DIAG_BOOL_PREVALENT_FINAL)

# 2. Load disease pattern mapping (Output from 01_hierarchical_clustering)
disease_patterns <- read_csv( file.path(PATH_INTERMEDIATE,"jaccard_disease_patterns.csv"))

head(disease_patterns)
```

# 2. Calculate Pattern Scores per Patient

We pivot the binary diagnosis matrix to a long format, join it with the assigned clusters, and summarize the total number of conditions each patient presents per pattern.

```{r compute-scores}
# Reshape diagnosis matrix and match with pattern mapping
patient_patterns_long <- diag_data %>%
  select(-vih_dt, -HIV_infection, -`HIV-related_disease`) %>% 
  pivot_longer(
    cols = -patient_id, 
    names_to = "enfermedad", 
    values_to = "has_disease"
  ) %>%
  filter(has_disease == 1) %>%
  inner_join(disease_patterns, by = "enfermedad")

# Compute count score and binary flag (present/absent) per patient and cluster
patient_scores_wide <- patient_patterns_long %>%
  group_by(patient_id, cluster) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = cluster, 
    values_from = count, 
    names_prefix = "pattern_",
    values_fill = 0
  )

# Ensure all original patients are retained (even if they have 0 conditions in all clusters)
all_patients <- diag_data %>% select(patient_id)

patient_pattern_scores <- all_patients %>%
  left_join(patient_scores_wide, by = "patient_id") %>%
  mutate(across(starts_with("pattern_"), ~ replace_na(.x, 0)))

# Add binary flags (1 if patient has >= 1 disease in that pattern, 0 otherwise)
patient_pattern_scores <- patient_pattern_scores %>%
  mutate(across(
    starts_with("pattern_"), 
    ~ ifelse(.x >= 1, 1, 0), 
    .names = "{.col}_bool"
  ))

head(patient_pattern_scores)

```

# 3. Summary Statistics of Pattern Scores

We inspect the prevalence and distribution of each comorbidity pattern across the entire cohort.

```{r summary-stats}
# Summary of disease counts per pattern
patient_pattern_scores %>% 
  select(starts_with("pattern_") & !ends_with("_bool")) %>% 
  summary()

# Prevalence of having at least 1 condition per pattern
patient_pattern_scores %>% 
  select(ends_with("_bool")) %>% 
  summarise(across(everything(), ~ mean(.x) * 100)) %>% 
  pivot_longer(cols = everything(), names_to = "Pattern", values_to = "Prevalence_pct")

```

# 4. Export Processed Scores

We save the resulting dataset containing patient IDs and their respective pattern scores for subsequent epidemiological validation against HIV status.

```{r export}
write_csv(patient_pattern_scores, file.path(PATH_INTERMEDIATE, "patient_pattern_scores.csv"))
```

