---
title: "Patient Pattern Scoring"
subtitle: "Mapping disease clusters onto individual patient profiles"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

```  
  
# 1. Introduction  
Once disease-level comorbidity patterns have been identified through hierarchical clustering, this script maps those disease patterns onto individual patients.

For each candidate clustering solution (k = 3 to 7), each patient receives a pattern-specific score based on the proportion of diseases defining that pattern that are present in the patient.

The resulting patient-level scores are retained for all candidate values of k. This allows the hierarchical clustering results to be evaluated at the same levels of granularity.  
  
A dominant pattern is also derived when one pattern has the highest score for a given patient and k. This is used only as a summary representation and does not imply that a patient belongs exclusively to a single comorbidity pattern.
  
# 2. Load Input Datasets

We import the necessary files:

1. The binary diagnosis matrix (`diagnosis_bool_prevalent_final.csv`).
2. The disease-cluster assignments obtained from hierarchical clustering
   for all candidate solutions (k = 3 to 7)
   (`jaccard_disease_patterns_multi_k.csv`).  
  
The clustering output determines which diseases belong to each pattern,
while the patient diagnosis matrix determines which of those diseases are
present in each individual

```{r load-data}
# 1. Load patient diagnostic matrix
diag_data <- read_csv(PATH_DIAG_BOOL_PREVALENT_FINAL)

# 2. Load disease-cluster assignments for all candidate k values
disease_patterns_multi <- read_csv(
  file.path(
    PATH_INTERMEDIATE,
    "jaccard_disease_patterns_multi_k.csv"
  )
) 
 
# Identify the disease variables used in hierarchical clustering
disease_vars <- disease_patterns_multi$enfermedad

# Check that every disease included in the clustering is also available
# in the patient-level diagnostic matrix
# The analysis is stopped if any disease is missing to prevent an
# incomplete or inconsistent patient-pattern mapping
missing_diseases <- setdiff(disease_vars, names(diag_data))

if (length(missing_diseases) > 0) {
  stop(
    "The following diseases from the clustering output are missing ",
    "from the diagnostic matrix: ",
    paste(missing_diseases, collapse = ", ")
  )
}

# Retain only the patient identifier and the disease variables used
# in the hierarchical clustering analysis
diag_data <- diag_data %>%
  select(patient_id, all_of(disease_vars))
```

# 3. Calculate Pattern Scores per Patient

The patient-level diagnostic matrix is converted to long format so that
each row represents one patient-disease combination.

The disease-to-cluster assignments are then joined to determine which
comorbidity pattern each disease belongs to for every candidate value of k.

For each patient, the number of diseases present within each pattern is counted. This count is subsequently divided by the total number of diseases defining that pattern, producing a proportional pattern score ranging from 0 to 1.

A score of 1 therefore indicates that the patient has all diseases defining that pattern, whereas a score of 0 indicates that none of the diseases defining the pattern are present.

```{r compute-scores}
# Define the candidate numbers of clusters evaluated in the
# hierarchical clustering analysis
k_candidates <- 3:7

# Expected column names containing the disease-to-cluster assignments
# for each candidate clustering solution
required_cluster_cols <- paste0("cluster_k", k_candidates)

# Validate that all required clustering solutions are available
missing_cluster_cols <- setdiff(
  required_cluster_cols,
  names(disease_patterns_multi)
)

if (length(missing_cluster_cols) > 0) {
  stop(
    "Missing clustering columns in disease pattern file: ",
    paste(missing_cluster_cols, collapse = ", ")
  )
}

# Ensure that each patient appears only once in the diagnostic matrix.
# Duplicate patient identifiers could otherwise lead to incorrect
# patient-level pattern scores
if (anyDuplicated(diag_data$patient_id) > 0) {
  stop("patient_id is not unique in the diagnostic matrix.")
}

# Convert the disease-to-cluster assignments from wide to long format
# Each row represents one disease assigned to one pattern for a given k
disease_patterns_long <- disease_patterns_multi %>%
  pivot_longer(
    cols = all_of(required_cluster_cols),
    names_to = "k",
    names_pattern = "cluster_k(\\d+)",
    values_to = "pattern"
  ) %>%
  mutate(
    k = as.integer(k),
    pattern = as.integer(pattern)
  )

# Calculate the number of diseases defining each pattern for each
# candidate value of k.
# This value is required to calculate the proportional pattern score
cluster_sizes <- disease_patterns_long %>%
  count(k, pattern, name = "n_diseases")

# Convert the patient-level diagnostic matrix to long format.
# Only diseases that are present are retained because absent diseases
# contribute zero to the pattern-specific count
patient_diagnoses_long <- diag_data %>%
  pivot_longer(
    cols = -patient_id,
    names_to = "enfermedad",
    values_to = "has_disease"
  ) %>%
  filter(has_disease == 1) %>%
  select(patient_id, enfermedad)

# Count the number of diseases present in each patient-pattern
# combination for every candidate value of k
patient_pattern_scores_long <- patient_diagnoses_long %>%
  inner_join(
    disease_patterns_long,
    by = "enfermedad",
    relationship = "many-to-many"
  ) %>%
  count(
    patient_id,
    k,
    pattern,
    name = "count"
  )

# Create complete patient x k x pattern grid
all_patient_patterns <- expand_grid(
  patient_id = diag_data$patient_id,
  k = k_candidates
) %>%
  left_join(
    disease_patterns_long %>%
      distinct(k, pattern),
    by = "k"
  )

# Add zero counts for patient-pattern combinations without any
# observed diseases and calculate the proportional pattern score
patient_pattern_scores_long <- all_patient_patterns %>%
  left_join(
    patient_pattern_scores_long,
    by = c("patient_id", "k", "pattern"),
    relationship = "one-to-one"
  ) %>%
  left_join(
    cluster_sizes,
    by = c("k", "pattern"),
    relationship = "many-to-one"
  ) %>%
  mutate(
    count = replace_na(count, 0L),
    proportion = count / n_diseases
  ) %>%
  arrange(k, patient_id, pattern)
```  
  
# 4. Pattern Representation and Dominant Pattern

For each patient, the proportional pattern score represents the proportion
of diseases defining a given pattern that are present in that patient. For example, if a pattern contains 10 diseases and a patient has 4 of
those diseases, the corresponding pattern score is 0.40.

This score ranges from 0 to 1 and is calculated independently for each
candidate number of clusters (k = 3 to 7).

A dominant pattern can optionally be derived as the pattern with the
highest proportional score within each k. This assignment is a derived
summary and does not imply that patients belong exclusively to a single
comorbidity pattern.
  
```{r}
# Determine the dominant pattern within each patient and
# candidate clustering solution
patient_pattern_scores_long <- patient_pattern_scores_long %>%
  group_by(patient_id, k) %>%
  mutate(
    # Highest proportional representation across patterns
    max_score = max(proportion, na.rm = TRUE),
    
    # Number of patterns sharing the maximum score
    n_max = sum(
      proportion == max_score & max_score > 0,
      na.rm = TRUE
    ),
    
    # Assign a dominant pattern only when the maximum is unique
    # and greater than zero
    dominant_pattern = case_when(
      max_score == 0 ~ NA_integer_,
      n_max > 1 ~ NA_integer_,
      TRUE ~ pattern[which.max(proportion)]
    )
  ) %>%
  ungroup()
```

# 5. Summary Statistics of Pattern Scores

We inspect the prevalence and distribution of each comorbidity pattern across the entire cohort.

```{r summary-stats}
# Distribution of proportional pattern scores
patient_pattern_scores_long %>%
  group_by(k, pattern) %>%
  summarise(
    # Number of patients with at least one disease belonging
    # to the corresponding pattern
    n_patients = sum(count > 0),
    
    # Mean and median proportional representation of the pattern
    # across the entire cohort
    mean_proportion = mean(proportion, na.rm = TRUE),
    median_proportion = median(proportion, na.rm = TRUE),
    
    # Standard deviation of the proportional scor
    sd_proportion = sd(proportion, na.rm = TRUE),
    
    # Percentage of the entire cohort with at least one disease
    # belonging to the corresponding pattern
    pattern_presence_pct = mean(count > 0) * 100,
    .groups = "drop"
  )
```

# 6. Export Processed Scores

Save the finalized dataset containing scores, proportions, and assigned clusters.

```{r export}
write_csv(patient_pattern_scores_long, file.path(PATH_INTERMEDIATE, "patient_pattern_scores_hierarchical_k3_k7.csv"))
```

