---
title: "Epidemiological Validation of Comorbidity Patterns"
subtitle: "Cross-sectional comparison between HIV cohort and matched controls"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)  
```  
  
# 1. Load Input Data  
   
We import:  
- Patient-level pattern scores (`patient_pattern_scores.csv`). 
- Matched cohort structure (`cohort_match_id.csv`).
- Demographic baseline variables (`bdu_full.csv`).  
  
```{r}
# Load datasets
scores_df <- read_csv(file.path(PATH_INTERMEDIATE, "patient_pattern_scores.csv"))
cohort_match <- read_csv(file.path(PATH_DATA,"cohort_match_id.csv"))
bdu_data <- read_csv(file.path(PATH_DATA,"bdu_full.csv"))

# Merge into a single master validation dataset
master_df <- cohort_match %>% 
  inner_join(scores_df, by = "patient_id") %>% 
  left_join(bdu_data %>% select(patient_id, vih_bool, edad, sexo, indice_privación), by = "patient_id")

head(master_df)
```

# 2. Descriptive Prevalence: HIV vs. Matched Controls  
  
We summarize the proportion of individuals presenting at least one condition ($\ge 1$) per comorbidity pattern, stratified by HIV status.  
  
```{r}
# Calculate pattern prevalence (binary flags) by HIV status
prevalence_summary <- master_df %>% 
  group_by(vih_bool) %>% 
  summarise(
    n_total = n(),
    across(ends_with("_bool"), ~ mean(.x) * 100, .names = "prev_{.col}")
  ) %>% 
  pivot_longer(
    cols = starts_with("prev_pattern_"),
    names_to = "Pattern",
    values_to = "Prevalence_pct"
  ) %>% 
  mutate(
    Pattern = str_replace(Pattern, "prev_pattern_", "Pattern "),
    Pattern = str_replace(Pattern, "_bool", ""),
    HIV_Status = if_else(vih_bool == 1, "HIV", "Control")
  )

# Display summary table
prevalence_summary %>% 
  select(HIV_Status, Pattern, Prevalence_pct) %>% 
  pivot_wider(names_from = HIV_Status, values_from = Prevalence_pct)
```
  
  
# 3. Univariate Statistical Testing  
  
We test for significant differences between HIV patients and controls:  
- McNemar's test for binary pattern flags (paired proportion comparison). 
- Wilcoxon signed-rank test for pattern disease counts (paired continuous burden).  
  
```{r}
# Identify pattern boolean columns
pattern_bool_cols <- names(master_df)[str_detect(names(master_df), "^pattern_\\d+_bool$")]

# Paired statistical comparisons (HIV vs matched control per pair)
# Reshape to paired format using match_id
paired_data <- master_df %>% 
  select(match_id, vih_bool, all_of(pattern_bool_cols)) %>% 
  pivot_wider(
    id_cols = match_id,
    names_from = vih_bool,
    values_from = all_of(pattern_bool_cols),
    names_prefix = "hiv_"
  )

# Function to run McNemar test per pattern
mcnemar_results <- map_dfr(pattern_bool_cols, function(col) {
  col_hiv <- paste0(col, "_hiv_1")
  col_ctrl <- paste0(col, "_hiv_0")
  
  tbl <- table(paired_data[[col_hiv]], paired_data[[col_ctrl]])
  
  # Run McNemar test if table dimensions allow
  mcn <- tryCatch(mcnemar.test(tbl), error = function(e) NULL)
  
  data.frame(
    Pattern = col,
    p_value = if (!is.null(mcn)) mcn$p.value else NA_real_
  )
})

mcnemar_results %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"))
```
  

# 4. Visualization: Odds Ratios per Pattern  
  
We fit a Conditional Logistic Regression (matching on match_id) to estimate Odds Ratios ($OR$) and $95\%$ Confidence Intervals for each comorbidity pattern.  
  
```{r}
library(survival) # Required for clogit (conditional logistic regression)

# Long-format estimation of conditional logistic regressions per pattern
or_results <- map_dfr(pattern_bool_cols, function(col_name) {
  formula_str <- paste0(col_name, " ~ vih_bool + strata(match_id)")
  fit <- clogit(as.formula(formula_str), data = master_df)
  
  res <- summary(fit)
  
  data.frame(
    Pattern = str_replace(col_name, "_bool", ""),
    OR = res$coefficients[1, "exp(coef)"],
    CI_lower = res$conf.int[1, "lower .95"],
    CI_upper = res$conf.int[1, "upper .95"],
    p_value = res$coefficients[1, "Pr(>|z|)"]
  )
})

# Forest plot of Odds Ratios
ggplot(or_results, aes(x = OR, y = Pattern)) +
  geom_point(size = 3, color = "#2c3e50") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, color = "#2c3e50") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Comorbidity Pattern Risk in HIV vs. Matched Controls",
    subtitle = "Conditional Logistic Regression (Adjusted by matching pair)",
    x = "Odds Ratio (log scale)",
    y = "Pattern Cluster"
  ) +
  theme_minimal()
```  
  
# 5. Export Validation Results  
We save the resulting statistical validation table.  
  
```{r}
write_csv(or_results, 
          file.path(PATH_RESULTS,"epidemiological_validation_or_results.csv"))
```  
