---
title: "Epidemiological Validation of Comorbidity Patterns"
subtitle: "Cross-sectional comparison between people living with HIV and matched controls"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)  
```  
  
# 1. Introduction

Once disease-level comorbidity patterns have been identified through
hierarchical clustering and mapped onto individual patients, this script
evaluates whether the resulting patterns differ between people living with
HIV (PLWH) and their matched controls.

The hierarchical clustering analysis provides candidate solutions with
k = 3 to k = 7 disease patterns. All candidate solutions are retained in
the epidemiological validation to allow comparison of the resulting
associations across different levels of clustering granularity.

For each value of k, two complementary measures of pattern representation
are evaluated:

1. Binary pattern presence, defined as having at least one disease belonging to the corresponding comorbidity pattern
2. Proportional pattern representation, defined as the proportion of diseases defining a pattern that are present in the individual

The matched study design is accounted for using conditional logistic
regression, with matching sets included as strata.  
  
# 2. Load Input Data and Build Complete Matched Cohort 
   
We import:  
- Patient-level pattern scores (`patient_pattern_scores_hierarchical_k3_k7.csv`). 
- Matched cohort structure 1:2 (`cohort_match_id.csv`).
- Demographic baseline variables (`bdu_full.csv`).  
  
```{r}
# Load datasets
scores_df    <- read_csv(file.path(PATH_INTERMEDIATE, "patient_pattern_scores_hierarchical_k3_k7.csv"))
cohort_match <- read_csv(file.path(PATH_DATA, "cohort_match_id.csv"))
bdu_data     <- read_csv(file.path(PATH_DATA, "bdu_full.csv"))

# Extract cases and controls separately and combine them into a
# single patient-level representation of the matched cohort
cases_df <- cohort_match %>% 
  select(match_id, patient_id) %>% 
  distinct()

controls_df <- cohort_match %>% 
  select(match_id, patient_id = control_id) %>% 
  distinct()

# Combine cases and controls while retaining their role within
# each matched set
long_cohort <- bind_rows(
  cases_df %>% mutate(role = "case"),
  controls_df %>% mutate(role = "control")
)

# Combine pattern scores with matching information, HIV status,
# and selected demographic variables
master_df <- long_cohort %>% 
  inner_join(scores_df, by = "patient_id") %>% 
  left_join(bdu_data %>% select(patient_id, vih_bool, edad, sexo, indice_privación), by = "patient_id")

head(master_df)
```

# 3. Descriptive Pattern Prevalence: PLWH vs Matched Controls

The matched cohort structure is first examined to verify the distribution
of cases and controls.

For each candidate clustering solution (k = 3 to k = 7), pattern presence
is defined as having at least one disease belonging to the corresponding
comorbidity pattern (`count > 0`).

The prevalence of each pattern is then calculated separately for PLWH and
matched controls. These descriptive estimates provide an initial
comparison before fitting conditional logistic regression models.  
  
```{r}
# Validate that the expected case/control structure is represented
# in the assembled matched cohort
cohort_counts <- master_df %>% 
  group_by(vih_bool) %>% 
  summarise(
    n_patients = n_distinct(patient_id),
    n_records = n()
  ) %>% 
  mutate(HIV_Status = if_else(as.logical(vih_bool), "PLWH (Cases)", "PNLWH (Matched Controls)"))

print(cohort_counts)

# Summarise pattern prevalence by HIV status.
# The dataset contains one row per patient, k, and pattern.
# Pattern presence is defined as having at least one disease
# belonging to the corresponding pattern (count > 0).
prevalence_summary <- master_df %>%
  group_by(k, pattern, vih_bool) %>%
  summarise(
    n_patients = n_distinct(patient_id),
    n_with_pattern = n_distinct(patient_id[count > 0]),
    prevalence_pct = mean(count > 0) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    HIV_Status = if_else(
      as.logical(vih_bool),
      "PLWH (Cases)",
      "PNLWH (Matched Controls)"
    ),
    Pattern = paste0("Pattern ", pattern)
  ) %>%
  select(
    k,
    HIV_Status,
    Pattern,
    n_patients,
    n_with_pattern,
    prevalence_pct
  )

print(prevalence_summary)
```
  
# 4. Conditional Logistic Regression: Binary Pattern Presence

Conditional logistic regression is used to estimate the association between HIV status and the presence of each comorbidity pattern.

For each candidate clustering solution and pattern, pattern presence is
defined as having at least one disease belonging to that pattern
(`count > 0`).

The models are stratified by matching set (`match_id`) to account for the
1:2 matched study design. The resulting odds ratios quantify the relative
odds of pattern presence among PLWH compared with their matched controls (PNLWH).
  
```{r}
or_results <- map_dfr(k_candidates, function(current_k) {
  
  patterns_in_k <- unique(master_df$pattern[master_df$k == current_k])
  
  map_dfr(patterns_in_k, function(p_idx) {
    
    # Create a binary indicator of pattern presence for the
    # current patient-pattern combination
    df_temp <- master_df %>%
      filter(k == current_k, pattern == p_idx) %>%
      mutate(pattern_bin = if_else(count > 0, 1, 0))
    
    # Require a minimum number of individuals with the pattern
    # before attempting model estimation
    if(sum(df_temp$pattern_bin == 1, na.rm = TRUE) < 5) return(NULL)
    
    fit <- tryCatch({
      clogit(vih_bool ~ pattern_bin + strata(match_id), data = df_temp, method = "efron")
    }, error = function(e) NULL)
    
    if(is.null(fit)) return(NULL)
    
    res <- summary(fit)
    
    data.frame(
      k        = current_k,
      Pattern  = p_idx,
      OR       = res$coefficients[1, "exp(coef)"],
      CI_lower = res$conf.int[1, "lower .95"],
      CI_upper = res$conf.int[1, "upper .95"],
      p_value  = res$coefficients[1, "Pr(>|z|)"]
    )
  })
})

# Adjust p-values across all tested k-pattern associations
# using the Benjamini-Hochberg false discovery rate procedure
or_results <- or_results %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

print(or_results)

# Forest Plot
ggplot(or_results, aes(x = OR, y = factor(Pattern), color = factor(k))) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Association Between Comorbidity Pattern Presence and HIV Status",
    subtitle = "Conditional logistic regression stratified by matching set",
    x = "Odds Ratio (log scale)",
    y = "Comorbidity Pattern",
    color = "Number of clusters (k)"
  ) +
  theme_minimal()
```  
# 5. Complementary Analysis: Proportional Pattern Representation  
  
For each patient and pattern, the score represents the proportion of
diseases defining that pattern that are present in the individual and
therefore ranges from 0 to 1.

The conditional logistic regression coefficient represents the change in
odds of being a PLWH associated with a one-unit increase in this score,
corresponding to a change from 0 to 1 in proportional pattern
representation.

This analysis complements the binary pattern-presence analysis by
considering the degree of representation of each pattern rather than only
whether at least one defining disease is present.  
  
```{r}
# Fit conditional logistic regression models using the proportional
# pattern score as a continuous predictor
score_or_results <- map_dfr(k_candidates, function(current_k) {
  
  patterns_in_k <- unique(master_df$pattern[master_df$k == current_k])
  
  map_dfr(patterns_in_k, function(p_idx) {
    
    df_temp <- master_df %>%
      filter(k == current_k, pattern == p_idx)
    
    fit <- tryCatch({
      clogit(vih_bool ~ proportion + strata(match_id), data = df_temp, method = "efron")
    }, error = function(e) NULL)
    
    if(is.null(fit)) return(NULL)
    
    res <- summary(fit)
    
    data.frame(
      k           = current_k,
      Pattern     = p_idx,
      OR_per_unit = res$coefficients[1, "exp(coef)"],
      CI_lower    = res$conf.int[1, "lower .95"],
      CI_upper    = res$conf.int[1, "upper .95"],
      p_value     = res$coefficients[1, "Pr(>|z|)"]
    )
  })
})

# Adjust p-values across all tested k-pattern associations
# using the Benjamini-Hochberg false discovery rate procedure
score_or_results <- score_or_results %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

print(score_or_results)
```  
  
# 6. Export Validation Results  
  
The results from both complementary analyses are exported:

1. Odds ratios for binary pattern presence
2. Odds ratios associated with continuous proportional pattern
   representation  
   
```{r}
write_csv(or_results, file.path(PATH_RESULTS, "epidemiological_validation_binary_or.csv"))
write_csv(score_or_results, file.path(PATH_RESULTS, "epidemiological_validation_score_or.csv"))
```  
