---
title: "Epidemiological Validation of Comorbidity Patterns"
subtitle: "Cross-sectional comparison between HIV cohort and matched controls"
author: "Inés García de la Peña Marco"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)  
```  
  
# 1. Introduction  
Once we have mapped the identifyied patterns to individual patient profiles in the previous steps, the next step is to validate these patterns.  
  
In this script, it is performed a comparative analysis between the HIV cohort (People Living with HIV, PLWH) and the cohort of non-HIV matched controls (NPLWH).  
  
By using paired statistical tests, as McNemar tests, and conditional logistic regression models adjusted for matching strata and standard errors, we evaluate if a specific pattern is over or underrepresented in PLWH compared to the control group (NPLWH).  
  
# 2. Load Input Data and Build Complete Matched Cohort 
   
We import:  
- Patient-level pattern scores (`patient_pattern_scores.csv`). 
- Matched cohort structure 1:2 (`cohort_match_id.csv`).
- Demographic baseline variables (`bdu_full.csv`).  
  
```{r}
# Load datasets
scores_df    <- read_csv(file.path(PATH_INTERMEDIATE, "patient_pattern_scores.csv"))
cohort_match <- read_csv(file.path(PATH_DATA, "cohort_match_id.csv"))
bdu_data     <- read_csv(file.path(PATH_DATA, "bdu_full.csv"))

# Extract cases and controls separately to unify them in a long format
cases_df <- cohort_match %>% 
  select(match_id, patient_id) %>% 
  distinct()

controls_df <- cohort_match %>% 
  select(match_id, patient_id = control_id) %>% 
  distinct()

# Bind cases and controls into a single matching dataframe
long_cohort <- bind_rows(cases_df, controls_df)

# Create master_df including pattern scores and demographics for all patients
master_df <- long_cohort %>% 
  inner_join(scores_df, by = "patient_id") %>% 
  left_join(bdu_data %>% select(patient_id, vih_bool, edad, sexo, indice_privación), by = "patient_id")

head(master_df)
```

# 3. Descriptive Prevalence: HIV vs Matched Controls  
  
We verify the matched cohort distribution ($1:2$ sampling structure) and calculate the proportion of individuals presenting at least one condition ($\ge 1$) per comorbidity pattern, stratified by HIV status.  
  
```{r}
# Validate that Cases (1) and Controls (0) are properly represented in the dataset
cohort_counts <- master_df %>% 
  group_by(vih_bool) %>% 
  summarise(
    n_patients = n_distinct(patient_id),
    n_records = n()
  ) %>% 
  mutate(HIV_Status = if_else(as.logical(vih_bool), "PLWH (Cases)", "PNLWH (Controls)"))

print(cohort_counts)

# Summary of pattern prevalence by group
prevalence_summary <- master_df %>% 
  group_by(vih_bool) %>% 
  summarise(
    n_total = n(),
    across(ends_with("_bool"), ~ mean(.x) * 100, .names = "prev_{.col}")
  ) %>% 
  pivot_longer(
    cols = starts_with("prev_pattern_"),
    names_to = "Pattern",
    values_to = "Prevalence_pct"
  ) %>% 
  mutate(
    Pattern = str_replace(Pattern, "prev_pattern_", "Pattern "),
    Pattern = str_replace(Pattern, "_bool", ""),
    HIV_Status = if_else(as.logical(vih_bool), "PLWH", "PNLWH")
  )

prevalence_summary %>% 
  select(HIV_Status, Pattern, Prevalence_pct) %>% 
  pivot_wider(names_from = HIV_Status, values_from = Prevalence_pct)
```
  
  
# 4. Univariate Statistical Testing (Paired Comparisons)  
  
We test for significant differences between PLWH cases and matched controls using McNemar's test for binary pattern flags.  
  
```{r}
# Identify pattern boolean columns
pattern_bool_cols <- names(master_df)[str_detect(names(master_df), "^pattern_\\d+_bool$")]

# Transform to wide format by match_id (Cases "1" vs Controls "0")
paired_data <- master_df %>% 
  mutate(vih_str = if_else(as.logical(vih_bool), "1", "0")) %>% 
  select(match_id, vih_str, all_of(pattern_bool_cols)) %>% 
  pivot_wider(
    id_cols = match_id,
    names_from = vih_str,
    values_from = all_of(pattern_bool_cols),
    names_prefix = "hiv_",
    values_fn = list
  ) %>% 
  mutate(across(-match_id, ~ map_dbl(.x, ~ if (length(.x) > 0) .x[1] else NA_real_)))

# Execute McNemar Test
mcnemar_results <- map_dfr(pattern_bool_cols, function(col) {
  col_hiv  <- paste0(col, "_hiv_1")
  col_ctrl <- paste0(col, "_hiv_0")
  
  if (!all(c(col_hiv, col_ctrl) %in% names(paired_data))) {
    return(data.frame(Pattern = str_replace(col, "_bool", ""), p_value = NA_real_))
  }
  
  tbl <- table(paired_data[[col_hiv]], paired_data[[col_ctrl]])
  mcn <- tryCatch(mcnemar.test(tbl), error = function(e) NULL)
  
  data.frame(
    Pattern = str_replace(col, "_bool", ""),
    p_value = if (!is.null(mcn)) mcn$p.value else NA_real_
  )
})

mcnemar_results <- mcnemar_results %>% 
  mutate(p_adj = p.adjust(p_value, method = "BH"))

print(mcnemar_results)
```
  

# 5. Multivariate Regression: Odds Ratios per Pattern  
  
We fit Conditional Logistic Regression models (`clogit`) stratified by matching set (`match_id`) with robust variance estimation (`cluster(patient_id)`) to account for $1:2$ control sampling with replacement.  
  
```{r}
# Long-format estimation of conditional logistic regressions per pattern
or_results <- map_dfr(pattern_bool_cols, function(col_name) {
  # Response variable is vih_bool and explanatory variable is the pattern flag
  formula_str <- paste0("vih_bool ~ ", col_name, " + strata(match_id)")
  
 # Fit conditional logistic regression with method = "efron" to allow robust cluster SEs
  fit <- clogit(
    as.formula(formula_str), 
    data = master_df, 
    cluster = patient_id,
    method = "efron"
  )
  
  res <- summary(fit)
  
  data.frame(
    Pattern  = str_replace(col_name, "_bool", ""),
    OR       = res$coefficients[1, "exp(coef)"],
    CI_lower = res$conf.int[1, "lower .95"],
    CI_upper = res$conf.int[1, "upper .95"],
    p_value  = res$coefficients[1, "Pr(>|z|)"]
  )
})

# Display numeric results
print(or_results)

# Forest plot of Odds Ratios
ggplot(or_results, aes(x = OR, y = Pattern)) +
  geom_point(size = 3, color = "#2c3e50") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, color = "#2c3e50") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_log10() +
  labs(
    title = "Comorbidity Pattern Risk in PLWH vs. Matched Controls",
    subtitle = "Conditional Logistic Regression stratified by match_id (Robust Cluster SE)",
    x = "Odds Ratio (log scale)",
    y = "Pattern Cluster"
  ) +
  theme_minimal()
```  
  
# 6. Export Validation Results  
We save the resulting statistical validation table:  
  
```{r}
write_csv(or_results, 
          file.path(PATH_RESULTS,"epidemiological_validation_or_results.csv"))
```  
